---
title: "Differential Expression: Likelihood Ratio Test model"
subtitle: "Fetal (fMSC) and adult (aMSC) osteogenic time courses, D0-D17"
output:
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
---

<!--
================================================================================
 WHAT THIS NOTEBOOK DOES
================================================================================
 Fits a DESeq2 likelihood ratio test per cohort, asking which genes show a
 Time:Condition interaction - i.e. genes whose behaviour over the time course
 differs between osteogenic induction (OD) and control (CTRL).

   full     ~ Donor + Time + Condition + Time:Condition
   reduced  ~ Donor + Time + Condition

 It then visualises the result three ways: PCA of the transformed counts
 (figure 2), clustered heatmaps of the LRT-significant genes with GO
 enrichment per cluster (figure 3), and paired fetal-vs-adult heatmaps of the
 top 50 genes from each cohort (figure 4).

 STRUCTURE
   1  Setup            libraries, output helpers, constants and palettes
   2  Helper functions every plot / table function, defined exactly once
   3  Data             import, protein-coding filter
   4  LRT model        DESeq2 per cohort, results exported
   5  Figure 2         PCA by donor, condition and time
   6  Figure 3         LRT-DEG heatmaps and per-cluster GO enrichment
   7  Figure 4         top-50 paired fetal | adult heatmaps
   8  Session info

 INPUTS   data/fMSC_norm_counts.csv, data/fMSC_meta.csv
          data/aMSC_norm_counts.csv, data/aMSC_meta.csv
 OUTPUTS  <OUTPUT_DIR>/*.svg, *.png, *.tsv, *.xlsx
          <OUTPUT_DIR>/patent/*  (greyscale versions for print)

 CONVENTION  every figure is produced by one function taking greyscale =
 TRUE/FALSE, so the colour and print versions cannot drift apart. This was the
 main source of divergence in the original notebook: the greyscale copies had
 been edited independently and no longer showed the same content as the colour
 ones (see the notes in sections 2.4 and 2.7).
================================================================================
-->

# 1 Setup

## 1.1 Libraries

Load order matters: annotation and Bioconductor packages first, tidyverse
**last**, so `dplyr::select` / `filter` are not masked by `AnnotationDbi::select`
or `biomaRt::select`.

```{r setup, message = FALSE, warning = FALSE, error = FALSE}
# --- project / environment -----------------------------------------------
library(renv)           # project library management
library(here)           # paths relative to the project root

# --- annotation and differential expression ------------------------------
library(DESeq2)         # LRT model, vst / varianceStabilizingTransformation
library(ashr)           # shrinkage estimator (available for lfcShrink)
library(org.Hs.eg.db)   # Ensembl -> symbol mapping
library(biomaRt)        # gene biotype lookup (protein-coding filter)
library(gprofiler2)     # GO enrichment per heatmap cluster

# --- plotting -------------------------------------------------------------
library(ComplexHeatmap) # clustered heatmaps
library(circlize)       # colorRamp2 colour scales
library(EnhancedVolcano)
library(egg)            # ggarrange for the 4-panel PCA strips
library(svglite)        # vector output

# --- data handling and export --------------------------------------------
library(openxlsx)       # write.xlsx
library(tidyverse)      # ggplot2, dplyr, tidyr - loaded LAST (see above)

# Removed relative to the original setup chunk:
#   library(gdata)      no gdata function is called anywhere in this notebook
#   library(circlize)   was loaded a second time inside the figure 3 chunk
#   library(ggplot2)    duplicate - part of tidyverse
```

## 1.2 Output helpers

```{r export-helpers}
# Single place to change where everything is written. Note the Pairwise
# Comparisons notebook may point somewhere else (e.g. "output_clean2"); set
# them to the same value if you want both notebooks in one folder.
OUTPUT_DIR <- "output"

# svglite() and png() do NOT create missing folders - they fail with
# "cannot open stream <path>". Create the destination first.
.ensure_dir <- function(path){
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  path
}

exportplots <- function(x, y, w, h){
  path <- .ensure_dir(here::here(OUTPUT_DIR, paste0(as.character(y), ".svg")))
  svglite(filename = path, width = w, height = h)
  on.exit(dev.off(), add = TRUE)   # close the device even if print(x) errors,
                                   # otherwise later plots land in this file
  print(x)
  message("wrote ", path)
  invisible(path)
}

# Vector PDF. cairo_pdf embeds fonts and handles transparency properly;
# the base pdf() fallback is only used if R was built without cairo.
exportplots_pdf <- function(x, y, w, h){
  path <- .ensure_dir(here::here(OUTPUT_DIR, paste0(as.character(y), ".pdf")))
  if (capabilities("cairo")) {
    cairo_pdf(filename = path, width = w, height = h)
  } else {
    pdf(file = path, width = w, height = h, useDingbats = FALSE)
  }
  on.exit(dev.off(), add = TRUE)
  print(x)
  message("wrote ", path)
  invisible(path)
}

# Raster fallback. Only worth using when a figure is too heavy to stay vector;
# 600 dpi is the usual journal minimum for a greyscale/line figure.
exportplots_png <- function(x, y, w, h, res = 600){
  path <- .ensure_dir(here::here(OUTPUT_DIR, paste0(as.character(y), ".png")))
  png(filename = path, width = w, height = h, units = "in", res = res)
  on.exit(dev.off(), add = TRUE)
  print(x)
  message("wrote ", path)
  invisible(path)
}

# Write the same figure in several formats in one call. Journals and patent
# offices rarely want the same thing, and re-running a heatmap is cheap.
exportplots_all <- function(x, y, w, h, formats = c("svg", "pdf")){
  writers <- list(svg = exportplots, pdf = exportplots_pdf, png = exportplots_png)
  invisible(vapply(formats, function(f) writers[[f]](x, y, w, h), character(1)))
}

dir.create(here::here(OUTPUT_DIR, "patent"), recursive = TRUE, showWarnings = FALSE)

# macOS reaps /var/folders/.../T/ on a timer. When it does, tempdir() keeps
# returning a path that no longer exists and anything writing a temp file
# fails - including ComplexHeatmap's rasteriser, which reports it as
#   unable to open .../RtmpXXXX/.heatmap_body_....png
# Recreating the directory fixes it without restarting R.
if (!dir.exists(tempdir())) dir.create(tempdir(), recursive = TRUE)
```

## 1.3 Constants and palettes

Every colour used in the notebook is named here, once, in a colour and a
greyscale variant. Nothing downstream hard-codes a colour.

```{r constants}
TIME_LEVELS <- c("D0", "D1", "D2", "D3", "D4", "D10", "D17")
COHORTS     <- c("fMSC", "aMSC")
PADJ_CUT    <- 0.01   # LRT significance threshold
N_TIME      <- 13     # columns per donor: D0 baseline + 6 OD + 6 CTRL

# --- PCA: points coloured by Time ----------------------------------------
TIME_COLS    <- c(D17 = "red",   D10 = "orange", D4 = "gray29", D3 = "gray44",
                  D2  = "gray59", D1 = "gray74", D0 = "gray89")
TIME_COLS_BW <- c(D17 = "black", D10 = "black",  D4 = "gray29", D3 = "gray44",
                  D2  = "gray59", D1 = "gray74", D0 = "gray89")
# In greyscale the two late timepoints are both black, so they are separated
# by shape instead: triangle for D17, square for D10, circle for the rest.
TIME_SHAPES  <- c(D17 = 17, D10 = 15, D4 = 16, D3 = 16, D2 = 16, D1 = 16, D0 = 16)

# --- heatmap annotations --------------------------------------------------
TIME_ANNO_COLS <- c(D17 = "gray1",  D10 = "gray15", D4 = "gray29", D3 = "gray44",
                    D2  = "gray59", D1  = "gray74", D0 = "gray89")  # already neutral
COND_COLS      <- c(Baseline = "green", OD = "magenta", CTRL = "lightblue")
COND_COLS_BW   <- c(Baseline = "white", OD = "grey40",  CTRL = "grey75")
ORIGIN_COLS    <- c(Fetal = "brown",  Adult = "forestgreen")
ORIGIN_COLS_BW <- c(Fetal = "grey30", Adult = "grey70")

# --- heatmap body ---------------------------------------------------------
# Greyscale cannot render a diverging palette, so low -> high reads as
# light -> dark and zero is read off the legend rather than seen as a flip.
# A symmetric limit that ignores the tails. Anchoring the scale at zero and
# saturating at the 98th percentile of |z| is what stops a handful of extreme
# genes from washing the rest of the heatmap into mid-tone. Pass z_limit
# explicitly to override, or set it to Inf for the old min-max behaviour.
z_auto_limit <- function(mat, q = 0.98){
  z <- stats::quantile(abs(mat), q, na.rm = TRUE)
  if (!is.finite(z) || z == 0) z <- max(abs(mat), na.rm = TRUE)
  unname(z)
}

# Greyscale: light = low, dark = high, mid-grey = zero. Monotonic, so it still
# reads correctly printed in black and white.
z_ramp_bw <- function(mat, z_limit = NULL){
  if (is.null(z_limit)) z_limit <- z_auto_limit(mat)
  if (!is.finite(z_limit)) {
    return(colorRamp2(c(min(mat, na.rm = TRUE), max(mat, na.rm = TRUE)),
                      c("white", "black")))
  }
  colorRamp2(c(-z_limit, 0, z_limit), c("white", "grey65", "black"))
}

# Colour: same limits as the greyscale ramp, so the two versions of a figure
# have identical contrast and only differ in hue.
z_ramp_col <- function(mat, z_limit = NULL){
  if (is.null(z_limit)) z_limit <- z_auto_limit(mat)
  if (!is.finite(z_limit)) z_limit <- max(abs(mat), na.rm = TRUE)
  colorRamp2(c(-z_limit, 0, z_limit), c("blue", "white", "red"))
}
```

# 2 Helper functions

Defined once each. In the original notebook every figure existed as two
hand-maintained copies, one colour and one greyscale, and they had drifted
apart; each function below takes `greyscale = TRUE/FALSE` instead.

## 2.1 PCA panel

One function covers all eight PCA panels: two cohorts by four groupings
(donor, condition, OD-only over time, CTRL-only over time).

```{r fn-plot-pca}
plot_pca <- function(vsd, group, title,
                     meta_df = NULL, condition_keep = NULL,
                     xlims = c(-30, 30), ylims = c(-30, 30),
                     greyscale = FALSE, point_size = 2){

  pcaData <- plotPCA(vsd, intgroup = group, returnData = TRUE)

  # Read the variance explained BEFORE any subsetting. `[.data.frame` drops
  # custom attributes, so calling attr(pcaData, "percentVar") after the subset
  # returns NULL and the axis labels silently become "PC1: NA% variance" -
  # which is what the original OD-only and CTRL-only panels did. The PCA is
  # fitted on all samples either way, so the full-data figure is the correct
  # number to report.
  percentVar <- round(100 * attr(pcaData, "percentVar"))

  # Optionally show only one arm of the experiment (OD or CTRL).
  if (!is.null(condition_keep)) {
    keep    <- rownames(meta_df)[meta_df$Condition %in% condition_keep]
    pcaData <- pcaData[rownames(pcaData) %in% keep, , drop = FALSE]
  }

  is_time <- identical(group, "Time")
  p <- ggplot(pcaData, aes(x = PC1, y = PC2))

  if (greyscale) {
    if (is_time) {
      # Colour ramp plus shape, because D17 and D10 are both black.
      p <- p +
        geom_point(aes(colour = .data[[group]], shape = .data[[group]]),
                   size = point_size) +
        scale_colour_manual(values = TIME_COLS_BW) +
        scale_shape_manual(values = TIME_SHAPES)
    } else {
      # Donor / Condition have no meaningful ordering, so shape alone.
      p <- p + geom_point(aes(shape = .data[[group]]),
                          size = point_size, colour = "black")
    }
  } else {
    p <- p + geom_point(aes(colour = .data[[group]]), size = point_size)
    if (is_time) p <- p + scale_colour_manual(values = TIME_COLS)
  }

  p +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() +
    theme_bw() +
    theme(aspect.ratio     = 1,
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position  = "bottom",
          legend.text      = element_text(size = 12),
          legend.title     = element_text(size = 14)) +
    ggtitle(title) +
    xlim(xlims[1], xlims[2]) +
    ylim(ylims[1], ylims[2]) +
    guides(colour = guide_legend(nrow = 1, byrow = TRUE),
           shape  = guide_legend(nrow = 1, byrow = TRUE))
}
```

## 2.2 Four-panel PCA strip for one cohort

```{r fn-pca-strip}
# Returns the four panels as a named list so they can be exported singly and
# as a strip, in either palette, without repeating the argument lists.
pca_panels <- function(vsd, meta_df, xlims, ylims, greyscale = FALSE){
  list(
    Donor     = plot_pca(vsd, "Donor",     "Donors",
                         xlims = xlims, ylims = ylims, greyscale = greyscale),
    Condition = plot_pca(vsd, "Condition", "Condition",
                         xlims = xlims, ylims = ylims, greyscale = greyscale),
    OD        = plot_pca(vsd, "Time",      "OD coloured by Time",
                         meta_df = meta_df, condition_keep = "OD",
                         xlims = xlims, ylims = ylims, greyscale = greyscale),
    CTRL      = plot_pca(vsd, "Time",      "CTRL coloured by Time",
                         meta_df = meta_df, condition_keep = "CTRL",
                         xlims = xlims, ylims = ylims, greyscale = greyscale)
  )
}

# Write the four panels individually plus the combined strip.
export_pca_panels <- function(panels, cohort, greyscale = FALSE){
  # "fMSC Donor PCA" / "patent/fMSC Donor PCA greyscale", and the 4-panel
  # strip as "fMSC PCA" with no panel name in the middle.
  stem <- function(nm = NULL) {
    base <- paste(c(cohort, nm, "PCA"), collapse = " ")
    if (greyscale) paste0("patent/", base, " greyscale") else base
  }
  for (nm in names(panels)) exportplots(panels[[nm]], stem(nm), h = 4, w = 8)
  exportplots(do.call(ggarrange, c(unname(panels), list(ncol = 4))),
              stem(), h = 4, w = 14)
  invisible(NULL)
}
```

## 2.3 Collapse donors to one column per timepoint

```{r fn-mean-by-timepoint}
# Sample columns are ordered donor-major: 13 timepoints repeated per donor.
# Averaging across donors means striding by N_TIME. The original hard-coded
# seq(i, 52, 13), which assumes exactly four donors; using ncol() keeps this
# correct if a cohort has a different number.
mean_by_timepoint <- function(mat, n_time = N_TIME){
  out <- matrix(NA_real_, nrow = nrow(mat), ncol = n_time,
                dimnames = list(rownames(mat),
                                sub(".*?_", "", colnames(mat)[seq_len(n_time)])))
  for (i in seq_len(n_time)) {
    out[, i] <- rowMeans(mat[, seq(i, ncol(mat), n_time), drop = FALSE])
  }
  # Columns arrive interleaved OD/CTRL; regroup as baseline, all OD, all CTRL.
  out <- out[, c(1, seq(2, n_time, 2), seq(3, n_time, 2)), drop = FALSE]
  colnames(out)[1] <- "D0"
  out
}
```

## 2.4 LRT-DEG heatmap (figure 3)

```{r fn-lrt-heatmap}
# The greyscale copy in the original dropped the Time / Condition annotation
# bar entirely, so it was not the same figure as the colour version. Here both
# palettes carry the annotation; only the colours differ.
lrt_heatmap <- function(mat, title, greyscale = FALSE,
                        km = 4, z_limit = NULL,
                        width_cm = 10, seed = 123,
                        use_raster = FALSE){

  time_labels <- factor(c("D0", rep(c("D1", "D2", "D3", "D4", "D10", "D17"), 2)),
                        levels = TIME_LEVELS)
  cond_labels <- c("Baseline", rep("OD", 6), rep("CTRL", 6))

  ha <- HeatmapAnnotation(
    Time      = time_labels,
    Condition = cond_labels,
    col       = list(Time      = TIME_ANNO_COLS,
                     Condition = if (greyscale) COND_COLS_BW else COND_COLS),
    annotation_name_side = "left")

  args <- list(matrix           = mat,
               show_column_dend = FALSE,
               show_row_names   = FALSE,
               show_row_dend    = FALSE,
               cluster_columns  = FALSE,
               top_annotation   = ha,
               name             = "Z-Score",
               column_title     = title,
               column_title_gp  = gpar(fontsize = 25, fontface = "bold"),
               width            = unit(width_cm, "cm"),
               row_km           = km,
               row_gap          = unit(2, "mm"),
               # ComplexHeatmap turns rasterising ON by itself above 2000
               # rows, then renders the body through a temporary PNG device.
               # That fails inside an open svglite device on some setups, and
               # whenever tempdir() has been reaped. These heatmaps are 13
               # columns wide, so drawing the cells as vectors is cheap and
               # keeps the SVG genuinely scalable. Set use_raster = TRUE (and
               # install ragg, raster_device = "agg_png") if the file size of
               # a very tall heatmap becomes a problem.
               use_raster       = use_raster)
  # Both palettes get an explicit ramp on the same limits. Previously only the
  # greyscale branch set one and the colour version fell back to
  # ComplexHeatmap's default, so the two were not directly comparable.
  args$col <- if (greyscale) z_ramp_bw(mat, z_limit) else z_ramp_col(mat, z_limit)

  set.seed(seed)   # row_km is k-means: the seed fixes the cluster assignment
  hm <- do.call(Heatmap, args)
  draw(hm, heatmap_legend_side = "left", annotation_legend_side = "left")
}
```

## 2.5 GO enrichment of heatmap clusters

```{r fn-cluster-go}
# Takes the *drawn* heatmap, because row_order() only exists after draw().
cluster_go <- function(hm, mat, gene_file, go_file){

  clusters <- row_order(hm)
  if (!is.list(clusters)) clusters <- list(`1` = clusters)  # km = 1 edge case

  clu_gene <- lapply(names(clusters), function(i){
    data.frame(Symbol  = rownames(mat)[clusters[[i]]],
               Cluster = paste0("cluster", i),
               stringsAsFactors = FALSE)
  })
  # Name the sheets. The original named the adult list but not the fetal one,
  # so the two workbooks had different sheet names for the same content.
  names(clu_gene) <- vapply(clu_gene, function(d) d$Cluster[1], character(1))
  write.xlsx(clu_gene, here(OUTPUT_DIR, gene_file), colNames = TRUE)

  clu_go <- lapply(clu_gene, function(d){
    res <- gost(query = d$Symbol, organism = "hsapiens",
                ordered_query = FALSE, multi_query = FALSE, significant = TRUE,
                exclude_iea = FALSE, measure_underrepresentation = FALSE,
                evcodes = FALSE, user_threshold = 0.05,
                correction_method = "g_SCS", domain_scope = "annotated",
                custom_bg = NULL, numeric_ns = "", sources = NULL,
                as_short_link = FALSE, highlight = TRUE)
    # gost() returns NULL when nothing is significant; write.xlsx cannot take
    # a NULL sheet, so substitute an empty frame.
    if (is.null(res) || is.null(res$result)) data.frame() else res$result
  })
  names(clu_go) <- names(clu_gene)
  write.xlsx(clu_go, here(OUTPUT_DIR, go_file), colNames = TRUE)

  invisible(list(genes = clu_gene, go = clu_go))
}
```

## 2.6 Paired fetal | adult matrix for the top-N genes

```{r fn-top-matrix}
# Builds the 26-column matrix behind figure 4: the same genes scored in both
# cohorts, fetal in columns 1:13 and adult in 14:26.
top_gene_matrix <- function(gene_ids, LRT, meta, n_time = N_TIME){

  per_cohort <- lapply(seq_along(LRT), function(i){
    mat <- assay(varianceStabilizingTransformation(LRT[[i]], blind = FALSE))
    goi <- mat[rownames(mat) %in% gene_ids, , drop = FALSE]
    sc  <- t(apply(goi, 1, scale))                 # z-score each gene
    colnames(sc) <- meta[[i]]$label
    sc <- sc[, !duplicated(sc[1, ]), drop = FALSE] # drop the repeated baseline
    mean_by_timepoint(sc, n_time)
  })

  # Keep only genes present in both cohorts, in a common order. The original
  # did this with two if/else blocks and a manual identical() check.
  common <- sort(intersect(rownames(per_cohort[[1]]), rownames(per_cohort[[2]])))
  m <- cbind(per_cohort[[1]][common, , drop = FALSE],
             per_cohort[[2]][common, , drop = FALSE])
  rownames(m) <- mapIds(org.Hs.eg.db, keys = common,
                        keytype = "ENSEMBL", column = "SYMBOL")
  m
}
```

## 2.7 Paired fetal | adult heatmap (figure 4)

```{r fn-paired-heatmap}
# Layout note: the colour original put one long title across a column-split
# heatmap, while the greyscale original titled each split and added an overall
# title. The second reads better, so both palettes now use it. Pass
# panel_titles = NULL for the original single-title colour layout.
paired_heatmap <- function(m, main_title, greyscale = FALSE,
                           z_limit = 2, width_cm = 10, seed = 123,
                           panel_titles = c("Baseline", "OD", "CTRL"),
                           use_raster = FALSE){

  time_labels <- factor(c("D0", rep(c("D1", "D2", "D3", "D4", "D10", "D17"), 2)),
                        levels = TIME_LEVELS)
  cond_labels <- factor(c("Baseline", rep("OD", 6), rep("CTRL", 6)),
                        levels = c("Baseline", "OD", "CTRL"))

  make_anno <- function(origin){
    HeatmapAnnotation(
      Time      = time_labels,
      Origin    = factor(rep(origin, N_TIME), levels = origin),
      Condition = cond_labels,
      col = list(Time      = TIME_ANNO_COLS,
                 Origin    = (if (greyscale) ORIGIN_COLS_BW else ORIGIN_COLS)[origin],
                 Condition = if (greyscale) COND_COLS_BW else COND_COLS),
      annotation_name_side = "left")
  }

  panel <- function(cols, origin, legend_name, show_legend){
    args <- list(matrix              = m[, cols],
                 name                = legend_name,
                 show_heatmap_legend = show_legend,
                 show_column_dend    = FALSE,
                 show_column_names   = TRUE,
                 show_row_names      = TRUE,
                 show_row_dend       = FALSE,
                 cluster_columns     = FALSE,
                 cluster_rows        = TRUE,
                 border              = TRUE,
                 top_annotation      = make_anno(origin),
                 column_title        = if (is.null(panel_titles)) main_title
                                       else panel_titles,
                 column_title_gp     = gpar(fontsize = 12),
                 width               = unit(width_cm, "cm"),
                 column_split        = cond_labels,
                 column_gap          = unit(5, "mm"),
                 row_gap             = unit(2, "mm"),
                 use_raster          = use_raster)
    if (greyscale) args$col <- z_ramp_bw(m, z_limit)
    do.call(Heatmap, args)
  }

  set.seed(seed)
  # Second panel shares the first panel's scale, so it needs a distinct
  # internal name and no legend of its own.
  combined <- panel(1:13, "Fetal", "Z-Score", TRUE) +
              panel(14:26, "Adult", "Z-Score_adult", FALSE)

  draw(combined,
       column_title        = main_title,
       column_title_gp     = gpar(fontsize = 20, fontface = "bold"),
       heatmap_legend_side = "left",
       annotation_legend_side = "left",
       ht_gap = unit(2, "cm"))
}
```

# 3 Data

## 3.1 Import counts and metadata

```{r load-data}
# `counts_raw`, not `counts`: DESeq2 exports a counts() accessor and having a
# list of the same name in the global environment is a trap waiting to spring.
counts_raw <- list(
  fMSC = read.csv(here::here("data", "fMSC_norm_counts.csv"), row.names = 1),
  aMSC = read.csv(here::here("data", "aMSC_norm_counts.csv"), row.names = 1))

meta <- list(
  fMSC = read.csv(here::here("data", "fMSC_meta.csv"), row.names = 1),
  aMSC = read.csv(here::here("data", "aMSC_meta.csv"), row.names = 1))

# Time must be an ordered factor or DESeq2 and every axis will sort D10 < D2.
meta <- lapply(meta, function(m){
  m$Time <- factor(m$Time, levels = TIME_LEVELS)
  m
})

# Sample labels used on the heatmaps: Donor_Condition_Time.
meta <- lapply(meta, function(m){
  m$label <- paste(m$Donor, m$Condition, m$Time, sep = "_")
  m
})

stopifnot(identical(names(counts_raw), names(meta)))
vapply(counts_raw, ncol, integer(1))   # samples per cohort
```

## 3.2 Protein-coding filter

```{r protein-coding}
mart <- useEnsembl(biomart = "genes",
                   dataset = "hsapiens_gene_ensembl",
                   mirror  = "useast")   # pinned mirror, safer for publication

coding.counts <- lapply(counts_raw, function(cts){
  annot <- getBM(attributes = c("ensembl_gene_id", "gene_biotype"),
                 filters    = "ensembl_gene_id",
                 values     = rownames(cts),
                 mart       = mart)
  pc <- annot$ensembl_gene_id[annot$gene_biotype == "protein_coding"]
  cts[rownames(cts) %in% pc, , drop = FALSE]
})

vapply(coding.counts, nrow, integer(1))   # protein-coding genes retained
```

# 4 The likelihood ratio test model

The interaction term is what the test is about: the reduced model keeps Time
and Condition as main effects, so a significant gene is one whose *time course*
differs between OD and CTRL, not merely one that differs between arms.

```{r lrt-model}
LRT     <- list()
LRT.res <- list()

for (cohort in COHORTS) {
  dds <- DESeqDataSetFromMatrix(
    countData = round(coding.counts[[cohort]]),
    colData   = meta[[cohort]],
    design    = ~ Donor + Time + Condition + Time:Condition)

  # Prefilter: at least 10 counts in 3 or more samples.
  dds <- dds[rowSums(counts(dds) >= 10) >= 3, ]

  dds <- DESeq(dds, test = "LRT", reduced = ~ Donor + Time + Condition)

  res <- results(dds)
  res$symbol    <- mapIds(org.Hs.eg.db, keys = rownames(res),
                          keytype = "ENSEMBL", column = "SYMBOL")
  res$EnsemblID <- rownames(res)

  LRT[[cohort]]     <- dds
  LRT.res[[cohort]] <- res
}

# Export the full results tables.
for (cohort in COHORTS) {
  out_file <- here::here(OUTPUT_DIR, paste0(cohort, ".LRT.res.tsv"))
  write.table(as.data.frame(LRT.res[[cohort]]), file = out_file,
              sep = "\t", row.names = FALSE)
  message("Saved file: ", out_file)
}

# Genes passing the LRT at padj < PADJ_CUT.
vapply(LRT.res, function(r) sum(r$padj < PADJ_CUT, na.rm = TRUE), integer(1))
```

# 5 Figure 2: principal component analysis

Transformed counts, `blind = FALSE`, so the design is used when estimating
dispersions. Panels a-d are fetal, e-h adult; each is produced in colour and in
greyscale from the same function.

```{r fig2-transform}
vsd <- lapply(LRT, function(dds) vst(dds, blind = FALSE))

# Axis ranges differ between cohorts, so they are named here rather than
# repeated in eight places.
PCA_LIMS <- list(
  fMSC = list(x = c(-30, 30), y = c(-30, 30)),
  aMSC = list(x = c(-30, 38), y = c(-25, 25)))
```

## 5.1 Fetal PCA (2a-2d)

```{r fig2-fetal}
f_pca <- pca_panels(vsd[["fMSC"]], meta[["fMSC"]],
                    PCA_LIMS$fMSC$x, PCA_LIMS$fMSC$y, greyscale = FALSE)
f_pca$Donor; f_pca$Condition; f_pca$OD; f_pca$CTRL
export_pca_panels(f_pca, "fMSC", greyscale = FALSE)

f_pca_bw <- pca_panels(vsd[["fMSC"]], meta[["fMSC"]],
                       PCA_LIMS$fMSC$x, PCA_LIMS$fMSC$y, greyscale = TRUE)
export_pca_panels(f_pca_bw, "fMSC", greyscale = TRUE)
```

## 5.2 Adult PCA (2e-2h)

```{r fig2-adult}
a_pca <- pca_panels(vsd[["aMSC"]], meta[["aMSC"]],
                    PCA_LIMS$aMSC$x, PCA_LIMS$aMSC$y, greyscale = FALSE)
a_pca$Donor; a_pca$Condition; a_pca$OD; a_pca$CTRL
export_pca_panels(a_pca, "aMSC", greyscale = FALSE)

a_pca_bw <- pca_panels(vsd[["aMSC"]], meta[["aMSC"]],
                       PCA_LIMS$aMSC$x, PCA_LIMS$aMSC$y, greyscale = TRUE)
export_pca_panels(a_pca_bw, "aMSC", greyscale = TRUE)
```

# 6 Figure 3: LRT-significant gene heatmaps

## 6.1 Build the scaled matrices

```{r fig3-matrices}
# Z-scored transformed counts, restricted to LRT-significant genes, with the
# repeated baseline columns collapsed and rows labelled by gene symbol.
hmap_mat <- lapply(COHORTS, function(cohort){
  sig <- rownames(LRT.res[[cohort]])[which(LRT.res[[cohort]]$padj < PADJ_CUT)]
  mat <- assay(varianceStabilizingTransformation(LRT[[cohort]], blind = FALSE))
  sig.mat <- mat[rownames(mat) %in% sig, , drop = FALSE]

  scaled <- t(apply(sig.mat, 1, scale))
  colnames(scaled) <- meta[[cohort]]$label
  clean <- scaled[, !duplicated(scaled[1, ]), drop = FALSE]
  rownames(clean) <- mapIds(org.Hs.eg.db, keys = rownames(clean),
                            keytype = "ENSEMBL", column = "SYMBOL")
  clean
})
names(hmap_mat) <- COHORTS

# One column per timepoint, averaged over donors.
mean_mat <- lapply(hmap_mat, mean_by_timepoint)

vapply(mean_mat, nrow, integer(1))   # LRT-significant genes plotted per cohort
```

## 6.2 Heatmaps, colour and greyscale (3b, 3c)

```{r fig3-heatmaps}
# The original built the adult greyscale heatmap from hmap_mat[[1]] - the
# FETAL matrix - and overwrote both mean.adult.donor and aMSC_LRT_Heatmap in
# the process, so the adult cluster GO downstream was run on fetal genes.
# Driving both cohorts from the same loop makes that class of mistake hard to
# reproduce.
lrt_hm <- list()

for (cohort in COHORTS) {
  hm <- lrt_heatmap(mean_mat[[cohort]], paste0(cohort, " LRT DEGs"),
                    greyscale = FALSE)
  exportplots(hm, paste0(cohort, "_LRT_Heatmap"), h = 4, w = 14)
  lrt_hm[[cohort]] <- hm   # keep the colour version: clusters come from it

  hm_bw <- lrt_heatmap(mean_mat[[cohort]], paste0(cohort, " LRT DEGs"),
                       greyscale = TRUE)
  exportplots_all(hm_bw, paste0("patent/3_", cohort, "_LRT_Heatmap greyscale"),
                  h = 4, w = 14, formats = c("svg", "pdf"))
}
```

## 6.3 GO enrichment per cluster

```{r fig3-cluster-go}
# Clusters are read off the colour heatmap. Both palettes use the same data
# and the same seed, so the k-means partition is identical either way.
# Filenames kept as in the original so downstream references still resolve.
GENE_FILES <- c(fMSC = "Fetal.LRT.ClusterGenes.xlsx",
                aMSC = "ad.LRT.ClusterGenes.xlsx")
GO_FILES   <- c(fMSC = "Fetal.LRT.GO.xlsx",
                aMSC = "ad.LRT.GO.xlsx")

clu <- list()
for (cohort in COHORTS) {
  clu[[cohort]] <- cluster_go(lrt_hm[[cohort]], mean_mat[[cohort]],
                              gene_file = GENE_FILES[[cohort]],
                              go_file   = GO_FILES[[cohort]])
}

vapply(clu, function(x) length(x$genes), integer(1))   # clusters per cohort
```

# 7 Figure 4: top-50 genes, fetal beside adult

The same 50 genes are shown in both cohorts, so the question is not "are these
genes significant" but "does the adult time course reproduce the fetal one".

```{r fig4-gene-lists}
top_lrt_ids <- function(res, n = 50){
  d <- data.frame(res)
  d <- d[which(d$padj < PADJ_CUT), ]
  d <- d[order(d$padj), ]
  head(d$EnsemblID, n)
}

f50_ids <- top_lrt_ids(LRT.res[["fMSC"]])
a50_ids <- top_lrt_ids(LRT.res[["aMSC"]])
```

## 7.1 Top 50 fetal genes (4a)

```{r fig4a}
f50_mat <- top_gene_matrix(f50_ids, LRT, meta)

# Titles say "most significant", not "upregulated": the LRT tests for an
# interaction and does not assign a direction.
exportplots(paired_heatmap(f50_mat,
                           "Top 50 most significant fetal LRT genes (fetal | adult)",
                           greyscale = FALSE),
            "Top 50 Fetal Upregulated genes hmap", h = 10, w = 15)

exportplots(paired_heatmap(f50_mat,
                           "Top 50 most significant fetal LRT genes (fetal | adult)",
                           greyscale = TRUE),
            "patent/Top 50 Fetal genes hmap greyscale", h = 10, w = 15)
```

## 7.2 Top 50 adult genes (4b)

```{r fig4b}
a50_mat <- top_gene_matrix(a50_ids, LRT, meta)

exportplots(paired_heatmap(a50_mat,
                           "Top 50 most significant adult LRT genes (fetal | adult)",
                           greyscale = FALSE),
            "Top 50 Adult Upregulated genes hmap", h = 10, w = 15)

exportplots(paired_heatmap(a50_mat,
                           "Top 50 most significant adult LRT genes (fetal | adult)",
                           greyscale = TRUE),
            "patent/Top 50 Adult genes hmap greyscale", h = 10, w = 15)
```

# 8 Session and reproducibility

```{r session-info}
sessionInfo()
```
