---
title: "Differential Expression: Likelihood Ratio Test Model"
output: html_notebook
---

# Setup Environment
```{r setup, error = FALSE, warning = FALSE}
#Load dependencies
library(renv)
library(ggplot2)
library(org.Hs.eg.db)
library(ComplexHeatmap)
library(EnhancedVolcano)
library(tidyverse)
library(gdata)
library(ashr)
library(openxlsx)
library(gprofiler2)
library(circlize)
library(egg)
library(biomaRt)
library(here)
library(DESeq2)

exportplots <- function(x, y, w, h){
  as.character(y)
  png(filename = here::here("output", paste0(y, ".png")), width = w, height = h, units = "in", res = 600)
  print(x)
  dev.off()
}


```

#Load in Data + Meta
```{r}
# Load fetal MSC data
f.counts <- read.csv(here::here("data", "fMSC_norm_counts.csv"), row.names = 1)
f.meta   <- read.csv(here::here("data", "fMSC_meta.csv"), row.names = 1)

# Load adult MSC data
a.counts <- read.csv(here::here("data", "aMSC_norm_counts.csv"), row.names = 1)
a.meta   <- read.csv(here::here("data", "aMSC_meta.csv"), row.names = 1)

#Create data list and meta list to hold both fetal and adult data
counts = list(f.counts, a.counts)
names(counts) = c("fMSC","aMSC")

meta = list(f.meta, a.meta)
names(meta) = c("fMSC","aMSC")

#Correct levels in meta file
meta[[1]]$Time <- factor(meta[[1]]$Time, levels = c("D0", "D1", "D2", "D3", "D4", "D10", "D17"))
meta[[2]]$Time <- factor(meta[[2]]$Time, levels = c("D0", "D1", "D2", "D3", "D4", "D10", "D17"))
```

#Filter for Protein Coding Genes
```{r}
mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  mirror = "useast"          # safer for publication
)
coding.counts = list()

for(i in 1:length(counts)){

  genes <- rownames(counts[[i]])
  annot <- getBM(
    attributes = c("ensembl_gene_id", "gene_biotype"),
    filters = "ensembl_gene_id",
    values = genes,
    mart = mart
    )
  
  # Keep protein-coding genes
  protein_coding_genes <- annot$ensembl_gene_id[annot$gene_biotype == "protein_coding"]
  coding.counts[[i]] <- counts[[i]][rownames(counts[[i]]) %in% protein_coding_genes, ]
}
```

#Generation of the Likelihood Ratio Test Model in DESEQ2
```{r}
LRT = list()
LRT.res = list()

for(i in 1:length(coding.counts)){
  #Create deseq dataset
  x = DESeqDataSetFromMatrix(countData = round(coding.counts[[i]]),
                                colData = meta[[i]],
                                design = ~ Donor + Time + Condition + Time:Condition)
  
  #prefilter for genes with less than 10 coding.counts in 3 or more samples
  keep = rowSums(counts(x) >= 10) >= 3
  x = x[keep,]
  
  #Construct the LRT model 
  x <- DESeq(x, test="LRT", reduced = ~ Donor + Time + Condition)
  x.res <- results(x)
  x.res$symbol <- mapIds(org.Hs.eg.db, keys = rownames(x.res), keytype = "ENSEMBL", column = "SYMBOL")
  x.res$EnsemblID <- rownames(x.res)
  
  #Output results to the pre-initialized lists
  LRT[[i]] = x
  LRT.res[[i]] = x.res
}

#Export LRT.res as xls for accessibility 
for(i in 1:length(LRT.res)){
  if (i == 1) {
    out_file <- here::here("output", "fMSC.LRT.res.tsv")
    write.table(as.data.frame(LRT.res[[i]]),
              file = out_file,
              sep = "\t",
              row.names = FALSE)
    message("Saved file: ", out_file)
    }
  else{
    out_file <- here::here("output", "aMSC.LRT.res.tsv")
    write.table(as.data.frame(LRT.res[[i]]),
              file = out_file,
              sep = "\t",
              row.names = FALSE)
    message("Saved file: ", out_file)
    }
}
```

#Plot PCA colouring Donor Differences Fig 2a
```{r}
#Generate transformed values from dds_F object for PCA 
vsd_F <- vst(LRT[[1]], blind = FALSE)

pcaData <- plotPCA(vsd_F, intgroup = c("Donor"),returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
p1 <- ggplot(pcaData, aes(PC1, PC2, color=Donor)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  theme_bw()+
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14))+  
  ggtitle("Donors") +
      xlim(-30, 30) + 
  ylim(-30, 30) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p1
```

#Plot PCA describing Condition Differences Fig 2b
```{r}
pcaData <- plotPCA(vsd_F, intgroup = c("Condition"),returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))


p2 <- ggplot(pcaData, aes(PC1, PC2, color=Condition)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  theme_bw()+
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14))+  
  ggtitle("Condition") +
        xlim(-30, 30) + 
  ylim(-30, 30) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p2
```

#Plot PCA showing Time differences but only for OD Fig 2c
```{r}
pcaData <- plotPCA(vsd_F, intgroup = c("Time"),returnData=TRUE)
pcaData <- pcaData[-(which(rownames(pcaData) %in% rownames(f.meta[which(f.meta$Condition == "CTRL"),]))),]
percentVar <- round(100 * attr(pcaData, "percentVar"))
p3 <- ggplot(pcaData, aes(PC1, PC2, color = Time)) +
  geom_point(size = 2) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom"
  ) +
  ggtitle("OD coloured by Time") +
  scale_color_manual(
    values = c(
      "D17" = "red",
      "D10" = "orange",
      "D4"  = "gray29",
      "D3"  = "gray44",
      "D2"  = "gray59",
      "D1"  = "gray74",
      "D0"  = "gray89"
    )
  ) +
        xlim(-30, 30) + 
  ylim(-30, 30) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p3
```

#Plot PCA showing Time differences but only for CTRL Fig 2d
```{r}
pcaData <- plotPCA(vsd_F, intgroup = c("Time"),returnData=TRUE)
pcaData <- pcaData[(which(rownames(pcaData) %in% rownames(f.meta[which(f.meta$Condition == "CTRL"),]))),]
percentVar <- round(100 * attr(pcaData, "percentVar"))
p4 <- ggplot(pcaData, aes(PC1, PC2, color=Time)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  theme_bw()+
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14))+  
  ggtitle("CTRL coloured by Time")+
  scale_color_manual(values = c("D17" = "red", "D10" = "orange", "D4" = "gray29", "D3" = "gray44",
                           "D2" = "gray59", "D1" = "gray74", "D0" = "gray89"))+
        xlim(-30, 30) + 
  ylim(-30, 30) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p4

exportplots(x = p1, y = "fMSC Donor PCA", h = 4, w=8)
exportplots(x = p2, y = "fMSC Condition PCA", h = 4, w=8)
exportplots(x = p3, y = "fMSC OD PCA", h = 4, w=8)
exportplots(x = p4, y = "fMSC CTRL PCA", h = 4, w=8)

exportplots(x = ggarrange(p1,p2,p3,p4, ncol = 4), y = "fMSC PCA", h = 4, w=14)  
```

#Plot PCA colouring Donor Differences (2e)
```{r}
#Generate transformed values from dds_F object for PCA 
vsd_A <- vst(LRT[[2]], blind = FALSE)

pcaData <- plotPCA(vsd_A, intgroup = c("Donor"),returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
p1 <- ggplot(pcaData, aes(PC1, PC2, color=Donor)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  theme_bw()+
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14))+  
  ggtitle("Donors") +
  xlim(-30, 38) + 
  ylim(-25, 25) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p1
```

#Plot PCA describing Condition Differences (2f)
```{r}
pcaData <- plotPCA(vsd_A, intgroup = c("Condition"),returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

p2 <- ggplot(pcaData, aes(PC1, PC2, color=Condition)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  theme_bw()+
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14))+  
  ggtitle("Condition") +
    xlim(-30, 38) + 
  ylim(-25, 25) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p2
```

#Plot PCA showing Time differences but only for OD (2g)
```{r}
pcaData <- plotPCA(vsd_A, intgroup = c("Time"),returnData=TRUE)
pcaData <- pcaData[-(which(rownames(pcaData) %in% rownames(a.meta[which(a.meta$Condition == "CTRL"),]))),]
percentVar <- round(100 * attr(pcaData, "percentVar"))
p3 <- ggplot(pcaData, aes(PC1, PC2, color = Time)) +
  geom_point(size = 2) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom"
  ) +
  ggtitle("OD coloured by Time") +
  scale_color_manual(
    values = c(
      "D17" = "red",
      "D10" = "orange",
      "D4"  = "gray29",
      "D3"  = "gray44",
      "D2"  = "gray59",
      "D1"  = "gray74",
      "D0"  = "gray89"
    ))+
    xlim(-30, 38) + 
  ylim(-25, 25) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p3
```

#Plot PCA showing Time differences but only for CTRL (2h)
```{r}
pcaData <- plotPCA(vsd_A, intgroup = c("Time"),returnData=TRUE)
pcaData <- pcaData[(which(rownames(pcaData) %in% rownames(a.meta[which(a.meta$Condition == "CTRL"),]))),]
percentVar <- round(100 * attr(pcaData, "percentVar"))
p4 <- ggplot(pcaData, aes(PC1, PC2, color=Time)) +
  geom_point(size=2) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()+
  theme_bw()+
  theme(
    aspect.ratio = 1,
    panel.grid.major = element_blank(),  # remove major grid
    panel.grid.minor = element_blank(),  # remove minor grid
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14))+  
  ggtitle("CTRL coloured by Time")+
  scale_color_manual(values = c("D17" = "red", "D10" = "orange", "D4" = "gray29", "D3" = "gray44",
                           "D2" = "gray59", "D1" = "gray74", "D0" = "gray89"))+
    xlim(-30, 38) + 
  ylim(-25, 25) +
  guides(color = guide_legend(nrow = 1, byrow = TRUE))

p4

exportplots(x = p1, y = "aMSC Donor PCA", h = 4, w=8)
exportplots(x = p2, y = "aMSC Condition PCA", h = 4, w=8)
exportplots(x = p3, y = "aMSC OD PCA", h = 4, w=8)
exportplots(x = p4, y = "aMSC CTRL PCA", h = 4, w=8)

exportplots(x = ggarrange(p1,p2,p3,p4, ncol = 4), y = "aMSC PCA", h = 4, w=14)  
```

#Scaled heatmaps from LRT DEGs (Fig 3)
```{r}
#Extract the Ensemble ID of significantly differential expressed according to the LRT model
sig.LRT = list()
for (i in 1:2){
  x = LRT.res[[i]][which(LRT.res[[i]]$padj < 0.01),]
  sig.LRT[[i]] = rownames(x)
}

#Generate better labels for visualization based off meta
for (i in 1:2) {
  meta[[i]]$label = paste(meta[[i]]$Donor,meta[[i]]$Condition, meta[[i]]$Time,sep="_")
}

#Generate a scaled matrix with LRT significant genes and transformed counts for plotting heatmaps
hmap_mat = list()
for (i in 1:2){
  x = varianceStabilizingTransformation(LRT[[i]], blind = FALSE) #Transform counts
  mat = assay(x)
  sig.mat = mat[which(rownames(mat) %in% sig.LRT[[i]]),] #Filter for LRT significant counts
  scaled.sig.mat = t(apply(sig.mat,1,scale)) #Scale the counts for plotting on heatmap
  colnames(scaled.sig.mat) = meta[[i]]$label #Change labels to labels column in meta
  clean.scaled.sig.mat = scaled.sig.mat[,duplicated(scaled.sig.mat[1,]) == FALSE] #Clean mat for duplicated baseline
  rownames(clean.scaled.sig.mat) = mapIds(org.Hs.eg.db, keys = rownames(clean.scaled.sig.mat), keytype = "ENSEMBL", column = "SYMBOL") #Replace with Gene Symbols for a more descriptive Heatmap

  hmap_mat[[i]] = clean.scaled.sig.mat
}
```

#Plot fMSC Heatmap (3b)
```{r}
mean.fetal.donor = matrix(NA, nrow = nrow(hmap_mat[[1]]), ncol = 13)
rownames(mean.fetal.donor) = rownames(hmap_mat[[1]])

for(i in 1:13){ #Loop to obtain the average expression for each gene between donors
  x = rowMeans(hmap_mat[[1]][,seq(i,52,13)])
  mean.fetal.donor[,i] = x
}

colnames(mean.fetal.donor) = sub(".*?_", "",colnames(hmap_mat[[1]][,1:13])) #Generate Labels for new matrix
mean.fetal.donor = mean.fetal.donor[,c(1, seq(2,13,2), seq(3,13,2))] #Reorder Columns to separate time and condition

Time_labels = c("D0", "D1","D2","D3","D4","D10", "D17","D1","D2","D3","D4","D10", "D17")
Time_labels = factor(Time_labels, levels = c("D0", "D1", "D2", "D3", "D4", "D10", "D17"))

ha = HeatmapAnnotation(Time = Time_labels, 
                       Condition = c("Baseline",                                                               "OD","OD","OD","OD","OD","OD","CTRL","CTRL","CTRL","CTRL","CTRL","CTRL"),
                       col = list(Time = c("D17" = "gray1", "D10" = "gray15", "D4" = "gray29", "D3" = "gray44",
                           "D2" = "gray59", "D1" = "gray74", "D0" = "gray89"),
                                  Condition = c("Baseline" = "green", "OD" = "magenta", "CTRL" = "lightblue" )),
                       annotation_name_side = "left"
                       )

#Correct the name of baseline to just D0
colnames(mean.fetal.donor)[1] <- "D0"

set.seed(123)
fMSC_LRT_Heatmap = Heatmap(mean.fetal.donor,
                     show_column_dend = FALSE,
                     show_row_names = F,
                     show_row_dend = FALSE,
                     cluster_columns = FALSE,
                     top_annotation = ha,
                     name = "Z-Score",
                     column_title = "fMSC LRT DEGs",
                     column_title_gp = gpar(fontsize = 25, fontface = "bold"),
                     width = unit(10,"cm"),
                     row_km = 4,
                     row_gap = unit(2,"mm")
                     )
fMSC_LRT_Heatmap = draw(fMSC_LRT_Heatmap, heatmap_legend_side = "left", annotation_legend_side = "left")
exportplots(x = fMSC_LRT_Heatmap, y = "fMSC_LRT_Heatmap", h = 4, w=14)  
```

#GO Enrichment fetal clusters from LRT heatmap (3b)
```{r}
#Extract each cluster and run enrichment.
fMSC.clusters = row_order(fMSC_LRT_Heatmap) #Extract clusters

f.clu.gene <- lapply(names(fMSC.clusters), function(i){

  symbols <- rownames(mean.fetal.donor)[fMSC.clusters[[i]]]

  out <- data.frame(
    Symbol  = symbols,
    Cluster = paste0("cluster", i),
    stringsAsFactors = FALSE
  )
  out
})


write.xlsx(f.clu.gene,"/output/Fetal.LRT.ClusterGenes.xlsx", colNames = TRUE) #Export Cluster Genes

f.clu.gsea.GO = list()
for(i in 1:length(f.clu.gene)){
  out.go <- gost(query = f.clu.gene[[i]][,1],
                 organism = "hsapiens", ordered_query = FALSE, 
                 multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                 measure_underrepresentation = FALSE, evcodes = FALSE, 
                 user_threshold = 0.05, correction_method = "g_SCS", 
                 domain_scope = "annotated", custom_bg = NULL, 
                 numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)
  f.clu.gsea.GO[[i]] <- out.go$result
}

a<-c() #Create a variable of cluster names the f.clu.gsea.GO indexes represent
for(i in 1:length(f.clu.gene)){
  new <- f.clu.gene[[i]][1,2]
  a <- c(a, new)
}

names(f.clu.gsea.GO) <- a #Assign meaningful element names to f.clu.gsea.GO
write.xlsx(f.clu.gsea.GO,"/output/Fetal.LRT.GO.xlsx", colNames = TRUE)
```

#Plot aMSC LRT-DEG Heatmap (3c)
```{r}
mean.adult.donor = matrix(NA, nrow = nrow(hmap_mat[[1]]), ncol = 13)
rownames(mean.adult.donor) = rownames(hmap_mat[[1]])

for(i in 1:13){ #Loop to obtain the average expression for each gene between donors
  x = rowMeans(hmap_mat[[1]][,seq(i,52,13)])
  mean.adult.donor[,i] = x
}

colnames(mean.adult.donor) = sub(".*?_", "",colnames(hmap_mat[[1]][,1:13])) #Generate Labels for new matrix
mean.adult.donor = mean.adult.donor[,c(1, seq(2,13,2), seq(3,13,2))] #Reorder Columns to separate time and condition

Time_labels = c("D0", "D1","D2","D3","D4","D10", "D17","D1","D2","D3","D4","D10", "D17")
Time_labels = factor(Time_labels, levels = c("D0", "D1", "D2", "D3", "D4", "D10", "D17"))

ha = HeatmapAnnotation(Time = Time_labels, 
                       Condition = c("Baseline",                                                               "OD","OD","OD","OD","OD","OD","CTRL","CTRL","CTRL","CTRL","CTRL","CTRL"),
                       col = list(Time = c("D17" = "gray1", "D10" = "gray15", "D4" = "gray29", "D3" = "gray44",
                           "D2" = "gray59", "D1" = "gray74", "D0" = "gray89"),
                                  Condition = c("Baseline" = "green", "OD" = "magenta", "CTRL" = "lightblue" )),
                       annotation_name_side = "left"
                       )

#Correct the name of baseline to just D0
colnames(mean.adult.donor)[1] <- "D0"

set.seed(123)
aMSC_LRT_Heatmap = Heatmap(mean.adult.donor,
                     show_column_dend = FALSE,
                     show_row_names = F,
                     show_row_dend = FALSE,
                     cluster_columns = FALSE,
                     top_annotation = ha,
                     name = "Z-Score",
                     column_title = "aMSC LRT DEGs",
                     column_title_gp = gpar(fontsize = 25, fontface = "bold"),
                     width = unit(10,"cm"),
                     row_km = 4, #Determined based off manual inspection
                     row_gap = unit(2,"mm")
                     )
aMSC_LRT_Heatmap = draw(aMSC_LRT_Heatmap, heatmap_legend_side = "left", annotation_legend_side = "left")
exportplots(x = aMSC_LRT_Heatmap, y = "aMSC_LRT_Heatmap", h = 4, w=14) 
```

#GO enchriment of adult clusters from LRT heatmap (3c)
```{r}
#Extract each cluster and run enrichment.
aMSC.clusters = row_order(aMSC_LRT_Heatmap) #Extract clusters

a.clu.gene <- lapply(names(aMSC.clusters), function(i){

  symbols <- rownames(mean.adult.donor)[aMSC.clusters[[i]]]

  out <- data.frame(
    Symbol  = symbols,
    Cluster = paste0("cluster", i),
    stringsAsFactors = FALSE
  )
  out
})

a<-c() #Create a variable of cluster names the a.clu.gsea.GO indexes represent
for(i in 1:length(a.clu.gene)){
  new <- a.clu.gene[[i]][1,2]
  a <- c(a, new)
}

names(a.clu.gene) <- a

write.xlsx(a.clu.gene,"/output/ad.LRT.ClusterGenes.xlsx", colNames = TRUE) #Export Cluster Genes

a.clu.gsea.GO = list()
for(i in 1:length(a.clu.gene)){
  out.go <- gost(query = a.clu.gene[[i]][,1],
                 organism = "hsapiens", ordered_query = FALSE, 
                 multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                 measure_underrepresentation = FALSE, evcodes = FALSE, 
                 user_threshold = 0.05, correction_method = "g_SCS", 
                 domain_scope = "annotated", custom_bg = NULL, 
                 numeric_ns = "", sources = NULL, as_short_link = FALSE, highlight = TRUE)
  a.clu.gsea.GO[[i]] <- out.go$result
}

a<-c() #Create a variable of cluster names the a.clu.gsea.GO indexes represent
for(i in 1:length(a.clu.gene)){
  new <- a.clu.gene[[i]][1,2]
  a <- c(a, new)
}

names(a.clu.gsea.GO) <- a #Assign meaningful element names to a.clu.gsea.GO
write.xlsx(a.clu.gsea.GO,"/output/ad.LRT.GO.xlsx", colNames = TRUE)
```

#Plot a heatmap of the top 50 most significantly DE genes in fMSC, and the corresponding expression in aMSC (4a)
```{r}
#Obtain top 50 LRT-DEG in aMSC during differentiation
f50 = data.frame(LRT.res[[1]])
f50 = f50[which(f50$padj < 0.01),]
f50 = f50[order(f50$padj),]
f50 = f50[1:50,]

hmap.mat = list()
for (i in 1:2){
  x = varianceStabilizingTransformation(LRT[[i]], blind = FALSE) #Transform counts
  mat = assay(x)
  GOIs = mat[which(rownames(mat) %in% f50$EnsemblID),] #Filter for Genes of interest
  scaled.hmap.mat = t(apply(GOIs,1,scale)) #Scale the counts for plotting on heatmap
  colnames(scaled.hmap.mat) = meta[[i]]$label #Change labels to labels column in meta
  clean.scaled.hmap.mat = scaled.hmap.mat[,duplicated(scaled.hmap.mat[1,]) == FALSE] #Clean mat for duplicated baseline
  #rownames(clean.scaled.sig.mat) = mapIds(org.Hs.eg.db, keys = rownames(clean.scaled.sig.mat), keytype = "ENSEMBL", column = "SYMBOL") #Replace with Gene Symbols for a more descriptive Heatmap
  
  hmap.mat[[i]] = clean.scaled.hmap.mat
}

#Calculate mean scaled count values
mean.hmap.mat = list()
for (i in 1:length(hmap.mat)){
  mean.mat = matrix(NA, nrow = nrow(hmap.mat[[i]]), ncol = 13) #Initialize a blank matrix to hold mean values
  colnames(mean.mat) = sub(".*?_", "",colnames(hmap.mat[[i]][,1:13])) #Generate Col Labels for new matrix
  rownames(mean.mat) = rownames(hmap.mat[[i]]) #Replacing with gene ID row names
  for(a in 1:13){ #Loop to obtain the average expression for each gene between donors
    av = rowMeans(hmap.mat[[i]][,seq(a, ncol(hmap.mat[[i]]) , 13)])
    mean.mat[,a] = av
  }
  mean.hmap.mat[[i]] = mean.mat
}

#fMSC dataset has 3 extra genes that the aMSC did not map to - these need to be removed
which(rownames(mean.hmap.mat[[1]]) %in% rownames(mean.hmap.mat[[2]]) == FALSE) #These are rows 26, 340 and 506
which(rownames(mean.hmap.mat[[2]]) %in% rownames(mean.hmap.mat[[1]]) == FALSE)

#Remove unmapped genes
if (length(which(rownames(mean.hmap.mat[[1]]) %in% rownames(mean.hmap.mat[[2]]) == FALSE)) > 0){
  f.mat = mean.hmap.mat[[1]][-c(which(rownames(mean.hmap.mat[[1]]) %in% rownames(mean.hmap.mat[[2]]) == FALSE)),]
}else{
  f.mat = mean.hmap.mat[[1]]
}
f.mat = f.mat[order(rownames(f.mat)),]
colnames(f.mat)[[1]] = "D0"

#Remove unmapped genes
if (length(which(rownames(mean.hmap.mat[[2]]) %in% rownames(mean.hmap.mat[[1]]) == FALSE)) > 0){
  a.mat = mean.hmap.mat[[2]][-c(which(rownames(mean.hmap.mat[[2]]) %in% rownames(mean.hmap.mat[[1]]) == FALSE)),]
} else{
    a.mat = mean.hmap.mat[[2]]
  }
a.mat = a.mat[order(rownames(a.mat)),]
colnames(a.mat)[[1]] = "D0"

identical(rownames(f.mat),rownames(a.mat)) #Check whether mat (fetal counts) and a.mat (adult counts) have the same order

f.mat <- cbind(f.mat[,-seq(3,13,2)], f.mat[,seq(3,13,2)]) #Group the IN columns together and the NIN columns together
colnames(f.mat)[1] <- "D0" #Change the name of the baseline column

a.mat <- cbind(a.mat[,-seq(3,13,2)], a.mat[,seq(3,13,2)]) #Group the IN columns together and the NIN columns together
colnames(a.mat)[1] <- "D0" #Change the name of the baseline column

a50.mat = cbind(f.mat, a.mat) #Combine the 2 data sets for the heatmap matrix 
rownames(a50.mat) = mapIds(org.Hs.eg.db, keys = rownames(a50.mat), keytype = "ENSEMBL", column = "SYMBOL") #Turn labels from ensembl id to symbols

#Plot the heatmap
set.seed(123)
Time_labels = factor(c("D0", 
                       rep(c("D1", "D2", "D3", "D4", "D10", "D17"),2)),
                     levels = c("D0", "D1", "D2", "D3", "D4", "D10", "D17"))
Fetal_label = factor(rep("Fetal",13), levels = "Fetal")
Adult_label = factor(rep("Adult",13), levels = "Adult")
Condition_Labels = factor(c("Baseline", rep("OD",6), rep("CTRL",6)), 
                       levels = c("Baseline","OD", "CTRL"))

f.ha = HeatmapAnnotation(Time = Time_labels,
                        Origin = Fetal_label,
                        Condition = Condition_Labels,
                        col = list(
                          Time = c("D17" = "gray1", "D10" = "gray15", "D4" = "gray29", "D3" = "gray44","D2" = "gray59", 
                                   "D1" = "gray74", "D0" = "gray90"),
                          Origin = c("Fetal" = "brown"),
                          Condition = c("Baseline" = "green", "OD" = "magenta","CTRL" = "lightblue")
                          ),
                        annotation_name_side = "left")

f.map = Heatmap(a50.mat[,1:13],
                     show_column_dend = FALSE,
                     show_row_names = T,
                     show_row_dend = FALSE,
                     cluster_columns = FALSE,
                cluster_rows = T,
                     top_annotation = f.ha,
                     name = "Z-Score",
                     column_title = "Top 50 Fetal Upregulated genes",
                     column_title_gp = gpar(fontsize = 25, fontface = "bold"),
                     width = unit(10,"cm"),
                     #row_km = 3,
                     column_split = Condition_Labels,
                     column_gap = unit(5,"mm"),
                     #use_raster = TRUE,
                     #raster_quality = 10,
                     row_gap = unit(2,"mm")
                     )

a.ha = HeatmapAnnotation(Time = Time_labels,
                        Origin = Adult_label,
                        Condition = Condition_Labels,
                        col = list(
                          Time = c("D17" = "gray1", "D10" = "gray15", "D4" = "gray29", "D3" = "gray44","D2" = "gray59", 
                                   "D1" = "gray74", "D0" = "gray90"),
                          Origin = c("Adult" = "forestgreen"),
                          Condition = c("Baseline" = "green", "OD" = "magenta","CTRL" = "lightblue")
                          ),
                        annotation_name_side = "left")

a.map = Heatmap(a50.mat[,14:26],
                     show_column_dend = FALSE,
                     show_row_names = T,
                     show_row_dend = FALSE,
                     cluster_columns = FALSE,
                cluster_rows = T,
                     top_annotation = a.ha,
                     name = "Z-Score",
                     column_title = "Top 50 Fetal Upregulated genes",
                     column_title_gp = gpar(fontsize = 25, fontface = "bold"),
                     width = unit(10,"cm"),
                     #row_km = 3,
                     column_split = Condition_Labels,
                     column_gap = unit(5,"mm"),
                     #use_raster = TRUE,
                     #raster_quality = 10,
                     row_gap = unit(2,"mm")
                     )

fc2.hmap = f.map + a.map

f.diff.genes.hmap = draw(fc2.hmap, heatmap_legend_side = "left", annotation_legend_side = "left", ht_gap = unit(2,"cm"))

exportplots(x = f.diff.genes.hmap,
            y = "Top 50 Fetal Upregulated genes hmap",
            h = 10,
            w = 15)

```

#4a but in greyscale
```{r}
## ---- greyscale Z-score ramp, defined ONCE and reused ---------------------
# Sequential: light = low, dark = high. Greyscale can't render a diverging
# palette, so zero is read off the legend rather than seen as a colour flip.
z_lim <- 2                                     # clip Z at +/- 2
z_col <- colorRamp2(c(-z_lim, 0, z_lim),
                    c("white", "grey65", "black"))

Condition_Labels = factor(c("Baseline", rep("OD", 6), rep("CTRL", 6)),
                          levels = c("Baseline", "OD", "CTRL"))

## ---- heatmaps ------------------------------------------------------------
set.seed(123)

f.map = Heatmap(a50.mat[, 1:13],
                col  = z_col,
                name = "Z-Score",
                show_column_dend  = FALSE,
                show_column_names = TRUE,
                show_row_names    = TRUE,
                show_row_dend     = FALSE,
                cluster_columns   = FALSE,
                cluster_rows      = TRUE,
                border            = TRUE,
                column_title      = c("Baseline", "OD", "CTRL"),
                column_title_gp   = gpar(fontsize = 12),
                width             = unit(10, "cm"),
                column_split      = Condition_Labels,
                column_gap        = unit(5, "mm"),
                row_gap           = unit(2, "mm"))

a.map = Heatmap(a50.mat[, 14:26],
                col  = z_col,                  # same mapping as fetal
                name = "Z-Score_adult",        # unique name
                show_heatmap_legend = FALSE,   # single shared legend
                show_column_dend  = FALSE,
                show_column_names = TRUE,
                show_row_names    = TRUE,
                show_row_dend     = FALSE,
                cluster_columns   = FALSE,
                cluster_rows      = TRUE,
                border            = TRUE,
                column_title      = c("Baseline", "OD", "CTRL"),
                column_title_gp   = gpar(fontsize = 12),
                width             = unit(10, "cm"),
                column_split      = Condition_Labels,
                column_gap        = unit(5, "mm"),
                row_gap           = unit(2, "mm"))

fc2.hmap = f.map + a.map

f.diff.genes.hmap = draw(fc2.hmap,
                         column_title    = "Top 50 most significant fetal LRT genes (fetal | adult)",
                         column_title_gp = gpar(fontsize = 20, fontface = "bold"),
                         heatmap_legend_side = "left",
                         ht_gap = unit(2, "cm"))

exportplots(x = f.diff.genes.hmap,
            y = "Top 50 Fetal genes hmap greyscale",
            h = 10,
            w = 15)

```

#Plot a heatmap of the top 50 most significantly DE genes in fMSC, and the corresponding expression in aMSC (4b)
```{r}
#Obtain top 50 LRT-DEG in aMSC during differentiation
a50 = data.frame(LRT.res[[2]])
a50 = a50[which(a50$padj < 0.01),]
a50 = a50[order(a50$padj),]
a50 = a50[1:50,]

GeneList = a50$EnsemblID

hmap.mat = list()
for (i in 1:2){
  x = varianceStabilizingTransformation(LRT[[i]], blind = FALSE) #Transform counts
  mat = assay(x)
  GOIs = mat[which(rownames(mat) %in% GeneList),] #Filter for Genes of interest
  scaled.hmap.mat = t(apply(GOIs,1,scale)) #Scale the counts for plotting on heatmap
  colnames(scaled.hmap.mat) = meta[[i]]$label #Change labels to labels column in meta
  clean.scaled.hmap.mat = scaled.hmap.mat[,duplicated(scaled.hmap.mat[1,]) == FALSE] #Clean mat for duplicated baseline
  #rownames(clean.scaled.sig.mat) = mapIds(org.Hs.eg.db, keys = rownames(clean.scaled.sig.mat), keytype = "ENSEMBL", column = "SYMBOL") #Replace with Gene Symbols for a more descriptive Heatmap
  
  hmap.mat[[i]] = clean.scaled.hmap.mat
}

#Calculate mean scaled count values
mean.hmap.mat = list()
for (i in 1:length(hmap.mat)){
  mean.mat = matrix(NA, nrow = nrow(hmap.mat[[i]]), ncol = 13) #Initialize a blank matrix to hold mean values
  colnames(mean.mat) = sub(".*?_", "",colnames(hmap.mat[[i]][,1:13])) #Generate Col Labels for new matrix
  rownames(mean.mat) = rownames(hmap.mat[[i]]) #Replacing with gene ID row names
  for(a in 1:13){ #Loop to obtain the average expression for each gene between donors
    av = rowMeans(hmap.mat[[i]][,seq(a, ncol(hmap.mat[[i]]) , 13)])
    mean.mat[,a] = av
  }
  mean.hmap.mat[[i]] = mean.mat
}

#fMSC dataset has 3 extra genes that the aMSC did not map to - these need to be removed
which(rownames(mean.hmap.mat[[1]]) %in% rownames(mean.hmap.mat[[2]]) == FALSE) #These are rows 26, 340 and 506
which(rownames(mean.hmap.mat[[2]]) %in% rownames(mean.hmap.mat[[1]]) == FALSE)

#Remove unmapped genes
if (length(which(rownames(mean.hmap.mat[[1]]) %in% rownames(mean.hmap.mat[[2]]) == FALSE)) > 0){
  f.mat = mean.hmap.mat[[1]][-c(which(rownames(mean.hmap.mat[[1]]) %in% rownames(mean.hmap.mat[[2]]) == FALSE)),]
}else{
  f.mat = mean.hmap.mat[[1]]
}
f.mat = f.mat[order(rownames(f.mat)),]
colnames(f.mat)[[1]] = "D0"

#Remove unmapped genes
if (length(which(rownames(mean.hmap.mat[[2]]) %in% rownames(mean.hmap.mat[[1]]) == FALSE)) > 0){
  a.mat = mean.hmap.mat[[2]][-c(which(rownames(mean.hmap.mat[[2]]) %in% rownames(mean.hmap.mat[[1]]) == FALSE)),]
} else{
    a.mat = mean.hmap.mat[[2]]
  }
a.mat = a.mat[order(rownames(a.mat)),]
colnames(a.mat)[[1]] = "D0"

identical(rownames(f.mat),rownames(a.mat)) #Check whether mat (fetal counts) and a.mat (adult counts) have the same order

f.mat <- cbind(f.mat[,-seq(3,13,2)], f.mat[,seq(3,13,2)]) #Group the IN columns together and the NIN columns together
colnames(f.mat)[1] <- "D0" #Change the name of the baseline column

a.mat <- cbind(a.mat[,-seq(3,13,2)], a.mat[,seq(3,13,2)]) #Group the IN columns together and the NIN columns together
colnames(a.mat)[1] <- "D0" #Change the name of the baseline column

a50.mat = cbind(f.mat, a.mat) #Combine the 2 data sets for the heatmap matrix 
rownames(a50.mat) = mapIds(org.Hs.eg.db, keys = rownames(a50.mat), keytype = "ENSEMBL", column = "SYMBOL") #Turn labels from ensembl id to symbols

#Plot the heatmap
set.seed(123)
Time_labels = factor(c("D0", 
                       rep(c("D1", "D2", "D3", "D4", "D10", "D17"),2)),
                     levels = c("D0", "D1", "D2", "D3", "D4", "D10", "D17"))
Fetal_label = factor(rep("Fetal",13), levels = "Fetal")
Adult_label = factor(rep("Adult",13), levels = "Adult")
Condition_Labels = factor(c("Baseline", rep("OD",6), rep("CTRL",6)), 
                       levels = c("Baseline","OD", "CTRL"))

f.ha = HeatmapAnnotation(Time = Time_labels,
                        Origin = Fetal_label,
                        Condition = Condition_Labels,
                        col = list(
                          Time = c("D17" = "gray1", "D10" = "gray15", "D4" = "gray29", "D3" = "gray44","D2" = "gray59", 
                                   "D1" = "gray74", "D0" = "gray90"),
                          Origin = c("Fetal" = "brown"),
                          Condition = c("Baseline" = "green", "OD" = "magenta","CTRL" = "lightblue")
                          ),
                        annotation_name_side = "left")

f.map = Heatmap(a50.mat[,1:13],
                     show_column_dend = FALSE,
                     show_row_names = T,
                     show_row_dend = FALSE,
                     cluster_columns = FALSE,
                     top_annotation = f.ha,
                     name = "Z-Score",
                     column_title = "Top 50 Adult Upregulated genes IN",
                     column_title_gp = gpar(fontsize = 25, fontface = "bold"),
                     width = unit(10,"cm"),
                     #row_km = 3,
                     column_split = Condition_Labels,
                     column_gap = unit(5,"mm"),
                     #use_raster = TRUE,
                     #raster_quality = 10,
                     row_gap = unit(2,"mm")
                     )

a.ha = HeatmapAnnotation(Time = Time_labels,
                        Origin = Adult_label,
                        Condition = Condition_Labels,
                        col = list(
                          Time = c("D17" = "gray1", "D10" = "gray15", "D4" = "gray29", "D3" = "gray44","D2" = "gray59", 
                                   "D1" = "gray74", "D0" = "gray90"),
                          Origin = c("Adult" = "forestgreen"),
                          Condition = c("Baseline" = "green", "OD" = "magenta","CTRL" = "lightblue")
                          ),
                        annotation_name_side = "left")

a.map = Heatmap(a50.mat[,14:26],
                     show_column_dend = FALSE,
                     show_row_names = T,
                     show_row_dend = FALSE,
                     cluster_columns = FALSE,
                     top_annotation = a.ha,
                     name = "Z-Score",
                     column_title = "Top 50 Adult Upregulated genes IN",
                     column_title_gp = gpar(fontsize = 25, fontface = "bold"),
                     width = unit(10,"cm"),
                     #row_km = 3,
                     column_split = Condition_Labels,
                     column_gap = unit(5,"mm"),
                     #use_raster = TRUE,
                     #raster_quality = 10,
                     row_gap = unit(2,"mm")
                     )

fc2.hmap = f.map + a.map

f.diff.genes.hmap = draw(fc2.hmap, heatmap_legend_side = "left", annotation_legend_side = "left", ht_gap = unit(2,"cm"))

exportplots(x = f.diff.genes.hmap,
            y = "Top 50 Adult Upregulated genes hmap",
            h = 10,
            w = 15)
```

#Session and Reproducibility
```{r}
sessionInfo()
```